Formal Verification of Real-Time Data Processing of the LHC Beam Loss Monitoring System: A Case Study

### 2011-05-28 fmics/paper.tex
Naghmeh's final changes from Dropbox

Under such a constraint, we have proved a preliminary result bounding the error of $\mathsf{output}\;D\;n\;x\;t$ compared to $\mathsf{exact}\;D\;n\;x\;t$.
\begin{theorem}\label{max_error}
For all $k$ and $D$ satisfying $|D\;(t'+1)-D\;t'|\leq k$ at all times, the following holds for all slices $n>0$, tap positions $x$, and times $t$:
\begin{multline*}
t>(\mathsf{tap}\;n\;x+1)\times(\mathsf{delay}\;n)+\mathsf{delay}\;n+\mathsf{delay\_sum}\;n\implies\\
|\mathsf{output}\;D\;n\;x\;t-\mathsf{exact}\;D\;n\;x\;t|\leq(\mathsf{tap}\;n\;x+1)\times(\mathsf{delay\;n})\times k\times(\mathsf{delay\_sum}\;n-1+t\operatorname{mod}\mathsf{delay}\;n)
\end{multline*}
Furthermore, for all $k$ there exists an input stream $D_k$ satisfying $|D_k\;(t'+1)-D_k\;t'|\leq k$ for all $t'$ and $|\mathsf{output}\;D_k\;n\;x\;t-\mathsf{exact}\;D_k\;n\;x\;t|=(\mathsf{tap}\;n\;x+1)\times(\mathsf{delay\;n})\times k\times(\mathsf{delay\_sum}\;n-1+t\operatorname{mod}\mathsf{delay}\;n)$ for all $n>0$, $x$, and $t>(\mathsf{tap}\;n\;x+1)\times(\mathsf{delay}\;n)+\mathsf{delay}\;n+\mathsf{delay\_sum}\;n$.
\end{theorem}
\begin{proof}
By applying Theorem~\ref{output_eq_exact}, we are left with an inequation involving the function $\mathsf{exact}$ but not $\mathsf{output}$, specifically between $\mathsf{exact}\;D\;n\;x\;t$ and $\mathsf{exact}\;D\;n\;x\;(\mathsf{last\_update}\;n\;t+1-\mathsf{delay\_sum}\;n)$.
Our assumption on $t$ ensures that we are never in the $0$ case of either Theorem~\ref{output_eq_exact} or of the definition of $\mathsf{exact}$.
The function $\mathsf{exact}$ is defined as a sum of consecutive values of $D$; we need to bound the difference between one such sum and another earlier one of the same size.
But if at each step the maximum difference is $k$, then the total difference is at most $k$ times the distance between the ends of the two sums, as required.
(We use the fact that $\mathsf{last\_update}\;n\;t=t-t\operatorname{mod}\mathsf{delay}\;n$.)
The input stream achieving this bound is given by $D_k\;t=k\times t$. \qed
\end{proof}

Theorem~\ref{max_error} gives a bound on the error of an SRS output at a given time in terms of the time step, the slice and tap numbers, and the assumed bound on the difference between consecutive counts.
This bound is tight for the conditions we assume, namely that the time step is sufficiently high and that the difference between consecutive counts is bounded, in the sense that it is achievable by an input stream satisfying those conditions.
To determine whether an error bound is safe, however, it turns out to be more useful to know the relative size of the error as a fraction of the true sum.
We have not yet characterized the relationship between our error bound at a given time and the true sum at that time.

@@

The use of mechanized theorem proving to verify the correctness of the SRS behavior complements an intensive verification effort based on simulation performed at CERN.
This effort targeted the validity of the SRS as an accurate and fast enough method by analyzing its behavior (a) in the boundaries of the threshold limits and (b) in expected types of beam losses, e.g., fast and steady losses.
Its results showed that the current implementation of the SRS, given in Figure~\ref{fig:srs}, satisfies the specification, that is, the running sums have a maximum 20\% relative error ($|\mathsf{RS}\;D\;n\;t - \mathsf{true\_sum}\;D\;n\;t| < 20\%\times(\mathsf{true\_sum}))$.

By contrast, the results in this paper apply to all possible input streams, not just the sample inputs considered at CERN.
However, the bound proved in Theorem~\ref{max_error} is not tight enough to satisfy the maximum 20\% relative error bound.
One of the potential reasons for this problem is not knowing how to characterize the input stream to the SRS.
Due to the physical nature of this problem, such a characterization of the input stream is not trivial.
While testing and simulation of the SRS on a limited set of input streams offers a level of confidence that the behavior of SRS satisfies this particular specification, constructing a formal proof for all possible behaviors requires a better understanding of the characteristics of the input stream.
One future research direction is to investigate whether constructing such a formal proof is feasible for the SRS component of BLMS.

>>>>

Under such a constraint, we have proved the follwoing result bounding the error of $\mathsf{output}\;D\;n\;x\;t$ compared to $\mathsf{exact}\;D\;n\;x\;t$.
\begin{theorem}\label{max_error}
For all $k$ and $D$ satisfying $|D\;(t'+1)-D\;t'|\leq k$ at all times, the following holds for all slices $n>0$, tap positions $x$, and times $t$:
\begin{multline*}
t>(\mathsf{tap}\;n\;x+1)\times(\mathsf{delay}\;n)+\mathsf{delay}\;n+\mathsf{delay\_sum}\;n\implies\\
|\mathsf{output}\;D\;n\;x\;t-\mathsf{exact}\;D\;n\;x\;t|\leq(\mathsf{tap}\;n\;x+1)\times(\mathsf{delay\;n})\times k\times\\
(\mathsf{delay\_sum}\;n-1+t\operatorname{mod}\mathsf{delay}\;n)
\end{multline*}
Furthermore, for all $k$ there exists an input stream $D_k$ satisfying $|D_k\;(t'+1)-D_k\;t'|\leq k$ for all $t'$ and 
\begin{multline*}
|\mathsf{output}\;D_k\;n\;x\;t-\mathsf{exact}\;D_k\;n\;x\;t|=(\mathsf{tap}\;n\;x+1)\times(\mathsf{delay\;n})\times k\times\\
(\mathsf{delay\_sum}\;n-1+t\operatorname{mod}\mathsf{delay}\;n)
\end{multline*}

for all $n>0$, $x$, and $t>(\mathsf{tap}\;n\;x+1)\times(\mathsf{delay}\;n)+\mathsf{delay}\;n+\mathsf{delay\_sum}\;n$.

\end{theorem}
\begin{proof}
By applying Theorem~\ref{output_eq_exact}, we are left with an inequation involving the function $\mathsf{exact}$ 
%but not $\mathsf{output}$
specifically between $\mathsf{exact}\;D\;n\;x\;t$ and $\mathsf{exact}\;D\;n\;x\;(\mathsf{last\_update}\;n\;t+1-\mathsf{delay\_sum}\;n)$.
Our assumption on $t$ ensures that we are never in the $0$ case of either Theorem~\ref{output_eq_exact} or of the definition of $\mathsf{exact}$.
The function $\mathsf{exact}$ is defined as a sum of consecutive values of $D$; we need to bound the difference between one such sum and another earlier one of the same size.
But if at each step the maximum difference is $k$, then the total difference is at most $k$ times the distance between the ends of the two sums, as required.
(We use the fact that $\mathsf{last\_update}\;n\;t=t-t\operatorname{mod}\mathsf{delay}\;n$.)
The input stream achieving this bound is given by $D_k\;t=k\times t$. \qed
\end{proof}

Theorem~\ref{max_error} gives a bound on the error of a running sum at a given time in terms of the time step, the slice and tap numbers, and the assumed bound on the difference between consecutive counts.
This bound is tight for the conditions we assume, namely that the time step is sufficiently high and that the difference between consecutive counts is bounded, in the sense that it is achievable by an input stream satisfying those conditions.
To determine whether an error bound is acceptable with respect to the specification of SRS, however, it turns out to be more useful to know the relative size of the error as a fraction of the true sum. 
According to the specification of SRS, the running sums should have a maximum 20\% relative error (($|\mathsf{RS}\;D\;n\;t - \mathsf{true\_sum}\;D\;n\;t| < 20\%\times(\mathsf{true\_sum}))$.
%To determine whether an error bound is safe, however, it turns out to be more useful to know the relative size of the error as a fraction of the true sum.
We have not yet characterized the relationship between our error bound at a given time and the true sum at that time.

@@

The use of mechanized theorem proving to verify the correctness of the SRS behavior complements an intensive verification effort based on simulation performed at CERN.
This effort targeted the validity of the SRS as an accurate and fast enough method by analyzing its behavior (a) in the boundaries of the threshold limits and (b) in expected types of beam losses, e.g., fast and steady losses.
Its results showed that the current implementation of the SRS, given in Figure~\ref{fig:srs}, satisfies the specification, that is, the running sums have a maximum 20\% relative error.

By contrast, the results in this paper apply to all possible input streams, not just the sample inputs considered at CERN -- our main contribution, as stated in Theorem 2, showed that the behavior of the SRS is correct for all possible input streams.
However, we do not know if the constraint on input stream given in Theorem 3 is sufficient to satisfy the maximum 20\% relative error bound.
%However, the bound proved in Theorem~\ref{max_error} is not tight enough to satisfy the maximum 20\% relative error bound.
One of the potential reasons for this problem is not knowing how to characterize the input stream to the SRS.
Due to the physical nature of this problem, such a characterization of the input stream is not trivial.
While testing and simulation of the SRS on a limited set of input streams offers a level of confidence that the behavior of SRS satisfies this particular specification, constructing a formal proof for all possible behaviors requires a better understanding of the characteristics of the input stream.
One future research direction is to investigate whether constructing such a formal proof is feasible for the SRS component of BLMS.
